\large{Vipassana for Hackers} \\ \Huge{Paper Four: Mastering Spacetime} \\ \large\textit{Version 0.1}

### 2019-12-26 paper-four/paper-four.tex
p4: formatting

With these restrictions in mind, anyone is of course welcome to read the
paper. However, much of what is being said will seem unnecessarily recondite to
someone without the requisite practical meditation experience.

@@

From one's very first Vipassana course --- in fact, from the discourse on the second
day --- a meditator is met with two messages: that of \textit{Eka Maggo} (the one and
only path out of suffering) and the idea that the Noble Eightfold Path is
``complete''. This section will address these two issues.

\subsection{Learning Vipassana Elsewhere}

For new meditators, Goenkaji addresses this issue during early
courses. Although it isn't a point that is repeatedly addressed, there are two quotes
worth highlighting in this regard:

@@

Dissecting these two ideas quickly brings us to the conclusion that Goenka's
instruction in no way implies that he is the only legitimate teacher of Vipassana ---
just the most accessible. Anyone who has ever attempted a 10-day self course will
attest to the sheer difficulty of attempting to recreate the environment of a
Vipassana Centre in one's own home. But it can be done and is even recommended in the
Day 11 discourse before the course ends. \todo{add reference to discourse}

Between the 10-day introductory course, the 3-day course, and the Satipatthana Sutta
course, the messaging becomes very clear: Vedana (sensations) are of utmost
importance and terminology is of no importance. Many techniques refer to themselves
as ``Vipassana'' but if they do not deal with sensation they are not relevant to the
discussion, as we will see below in \textit{How is Vipassana ``Complete''?} Other
techniques, such as Zazen and Kundalini, which deal with sensation at least in part
are of interest to a meditator convinced of the value of sensation analysis. It is up
to each meditator to decide whether they prefer one technique or the other for
analyzing sensation and pursue that one technique vigorously. The reasons for
choosing only a single technique will be discussed in \textit{Warnings: Do not mix
  meditations}.

@@

\subsection{How is Vipassana ``Complete''?}

@@

@@

\subsection{Mastering Spacetime}

@@

\subsubsection{Anapana}

@@

\subsubsection{Vipassana}

@@

\subsubsection{Continuity}

@@

\section{structure: the requirement that meditation be taught for free (expand)}

@@

\begin{itemize}
  \item This document itself is a hindrance
  \item Don't get bogged down by ``perfect objectivity'' or ``perfect equanimity'' - yes muscles and blood flow follow your attention
  \item Do not use meditation to get high
\end{itemize}

\subsection{Do not mix meditations}

>>>>

With these restrictions in mind, anyone is of course welcome to read the
paper. However, much of what is being said will seem unnecessarily recondite to
someone without the requisite practical meditation experience. For meditators and
non-meditators, both, it is recommended that one read the \textit{Warnings} section
--- in particular, \textit{This document is a hindrance} --- before preceding
further.

@@

From one's very first Vipassana course --- in fact, from the discourse on the second
day --- a meditator is met with two messages: that of \textit{Eka Maggo} (the one and
only path out of suffering) and the idea that the Noble Eightfold Path is
``complete''. The following sections will address these two issues.

\section{Learning Vipassana Elsewhere}

For new meditators, S.N. Goenka addresses this issue during early
courses. Although it isn't a point that is repeatedly addressed, there are two quotes
worth highlighting in this regard:

@@

Dissecting these two ideas quickly brings us to the conclusion that Goenka's
instruction in no way implies that he is the only legitimate teacher of Vipassana ---
he is simply the most accessible. Anyone who has ever attempted a 10-day self course will
attest to the sheer difficulty of attempting to recreate the environment of a
Vipassana Centre in one's own home. But it can be done and is even recommended in the
Day 11 discourse before the course ends. \todo{add reference to discourse}

Between the 10-day introductory course, the 3-day course, and the Satipatthana Sutta
course, the messaging becomes very clear: Vedana \todo{add ref to vedana} (sensations) are of utmost
importance and terminology is of no importance. Many techniques refer to themselves
as ``Vipassana'' but if they do not deal with sensation they are not relevant to the
discussion, as we will see below in \textit{How is Vipassana ``Complete''?} Other
techniques, such as Zazen and Kundalini, which deal with sensation at least in part
are of interest to a meditator convinced of the value of sensation analysis. It is up
to each meditator to decide whether they prefer one technique or the other for
analyzing sensation and pursue that one technique vigorously. The reasons for
choosing only a single technique will be discussed in \textit{Warnings: Do not mix
  meditations}.

@@

\section{How is Vipassana ``Complete''?}

@@

\textbf{Kundalini} is the practice of repeatedly examining the chakras, points
largely located along the spinal cord. This is actually the final location of
attention in Vipassana, once the body-skin and internal body are dissolved to the
meditator's awareness. Kundalini can be thought of as a very specific subset of
Vipassana.  \todo{add ref to q \& a}

@@

\section{Mastering Spacetime}

@@

\subsection{Anapana}

@@

\subsection{Vipassana}

@@

\subsection{Continuity}

@@

\section{Structure: Cost}

I expect that many readers of this paper will be using it to triangulate their
opinion on their personal choice of lifetime meditation technique. I hope that it is
helpful in that regard and I would like to add one point, orthogonal to technique, to
that matter: cost.

Meditation, whatever technique you ultimately choose (and it is strongly advised that
you choose only one: see \textit{Do not mix techniques} below), must be taught for
free. The teacher, his or her organization, staff or servers, the facilities --- this
must all be made available for free.

Commercial meditations are always going to cause problems for the meditator for the
simple reason that, while one may progress a great deal in one's own meditation while
paying for it, the question of universality will arise in the mind of the meditator,
sooner or later. \textit{Can everyone practice this technique? As I am learning it?}
If not, the technique cannot be considered universal. If the question of universality
has not yet arisen in your mind, trust that it will. A meditation capable of
liberating only an individual, or only those who can afford it, is not much of a
meditation at all. The teaching is unlikely to have longevity, to persist, and the
value is unlikely to emerge as a global phenonmenon.

The idea of \textit{Eka Maggo}, the one and only path, may be interpreted any number
of ways. Some interpretations may even be paradoxical or inherently
self-referential. However, the very concept of a single path implies that a technique
is both \textit{complete} and \textit{universal}. It is strongly suggested you
repeatedly apply this test to any technique you are considering for lifetime
practice.

@@

\begin{itemize}
  \item This document is a hindrance
  \item Don't get bogged down by ``perfect objectivity'' or ``perfect equanimity'' - yes muscles and blood flow follow your attention
  \item Do not use meditation to get high
\end{itemize}

\subsection{Do not mix techniques}
